---
title: "midterm2_sheet"
output: 
  html_document: 
    keep_md: true
date: "2024-02-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## `pivot_longer()`  
Scientists frequently use spreadsheets that are organized to make data entry efficient. This is often referred to as **wide format**. Unfortunately, the wide format creates a problem because column names may actually represent values of a variable. The command `pivot_longer()` shifts data from wide to long format.   

Rules:  
+ `pivot_longer`(cols, names_to, values_to)
+ `cols` - Columns to pivot to longer format
+ `names_to` - Name of the new column; it will contain the column names of gathered columns as values
+ `values_to` - Name of the new column; it will contain the data stored in the values of gathered columns

## `pivot_wider()`
The opposite of `pivot_longer()`. You use `pivot_wider()` when you have an observation scattered across multiple rows. In the example below, `cases` and `population` represent variable names not observations.  

Rules:  
+ `pivot_wider`(names_from, values_from)  
+ `names_from` - Values in the `names_from` column will become new column names  
+ `values_from` - Cell values will be taken from the `values_from` column  

## Libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(naniar)
library(janitor)
```

## Grammar of Graphics
The ability to quickly produce and edit graphs and charts is a strength of R. These data visualizations are produced by the package `ggplot2` and it is a core part of the tidyverse. The syntax for using ggplot is specific and common to all of the plots. This is what Hadley Wickham calls a [Grammar of Graphics](http://vita.had.co.nz/papers/layered-grammar.pdf). The "gg" in `ggplot` stands for grammar of graphics.

## Philosophy
What makes a good chart? In my opinion a good chart is elegant in its simplicity. It provides a clean, clear visual of the data without being overwhelming to the reader. This can be hard to do and takes some careful thinking. Always keep in mind that the reader will almost never know the data as well as you do so you need to be mindful about presenting the facts.  

## Data Types
We first need to define some of the data types we will use to build plots.  

+ `discrete` quantitative data that only contains integers
+ `continuous` quantitative data that can take any numerical value
+ `categorical` qualitative data that can take on a limited number of values

## Basics
The syntax used by ggplot takes some practice to get used to, especially for customizing plots, but the basic elements are the same. It is helpful to think of plots as being built up in layers.  

In short, **plot= data + geom_ + aesthetics**.  

We start by calling the ggplot function, identifying the data, and specifying the axes. We then add the `geom` type to describe how we want our data represented. Each `geom_` works with specific types of data and R is capable of building plots of single variables, multiple variables, and even maps. Lastly, we add aesthetics.

## Example
To make things easy, let's start with some built in data.
```{r}
names(iris)
```

```{r}
glimpse(iris)
```

To make a plot, we need to first specify the data and map the aesthetics. The aesthetics include how each variable in our data set will be used. In the example below, I am using the aes() function to identify the x and y variables in the plot.
```{r}
ggplot(data=iris, #specify the data
       mapping=aes(x=Species, y=Petal.Length)) #map the aesthetics
```

Notice that we have a nice background, labeled axes, and even a value range of our variables on the y-axis- but no plot. This is because we need to tell ggplot how we want our data represented. This is called the geometry or `geom()`. There are many types of `geom`, see the ggplot [cheatsheet](https://posit.co/resources/cheatsheets/).

Here we specify that we want a boxplot, indicated by `geom_boxplot()`.
```{r}
ggplot(data=iris, #specify the data
       mapping=aes(x=Species, y=Petal.Length))+ #map the aesthetics
  geom_boxplot() #add the plot type
```

## Practice
1. Use the iris data to build a scatterplot that compares sepal length vs. sepal width. Use the cheat sheet for help to find the correct `geom_` for a scatterplot.
```{r}
ggplot(data=iris, 
       mapping=aes(x=Sepal.Width, y=Sepal.Length))+ 
  geom_point() 
```

## Scatterplots and barplots
Now that we have a general idea of the syntax, let's start by working with two common plots: 1) scatter plots and 2) bar plots.

## Data
**Database of vertebrate home range sizes.**  
Reference: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
Data: http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1  
```{r}
homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")
```

## Practice
1. What is the structure of the `homerange` data? Does it have any NA's? Is it tidy? Do a quick exploratory analysis of your choice below.

### 1. Scatter Plots
Scatter plots are good at revealing relationships that are not readily visible in the raw data. For now, we will not add regression aka. "best of fit" lines or calculate any r^2^ values.  

In the case below, we are exploring whether or not there is a relationship between animal mass and home range. We are using the **log transformed values** because there is a large difference in mass and home range among the different species in the data.
```{r}
names(homerange)
```

```{r}
ggplot(data=homerange, #specify the data
       mapping=aes(x=log10.mass, y=log10.hra))+ #map the aesthetics
  geom_point() #add the plot type
```

In big data sets with lots of overlapping values, over-plotting can be an issue. `geom_jitter()` is similar to `geom_point()` but it helps with over plotting by adding some random noise to the data and separating some of the individual points.
```{r}
ggplot(data=homerange, mapping=aes(x=log10.mass, y=log10.hra))+
  geom_jitter()
```

To add a regression (best of fit) line, we just add another layer.
```{r}
ggplot(data=homerange, mapping=aes(x=log10.mass, y=log10.hra))+
  geom_point()+
  geom_smooth(method=lm, se=T) #add a regression line
```

### Practice
1. What is the relationship between log10.hra and log10.preymass? What do you notice about how ggplot treats NA's?
```{r}
ggplot(data=homerange, mapping=aes(x=log10.hra, y=log10.preymass))+
  geom_jitter(na.rm=T)+
  geom_smooth(method=lm, se=T)
```

### Bar Plot: `geom_bar()`
The simplest type of bar plot counts the number of observations in a categorical variable. In this case, we want to know how many observations are present in the variable `trophic.guild`. Notice that we do not specify a y-axis because it is count by default.  

```{r}
names(homerange)
```

```{r}
homerange %>% 
  count(trophic.guild)
```

Also notice that we can use pipes! The `mapping=` function is implied by `aes` and so is often left out. 
```{r}
homerange %>% 
  ggplot(aes(x=trophic.guild)) + 
  geom_bar() #good for counts
```

### Bar Plot: `geom_col()`
Unlike `geom_bar()`, `geom_col()` allows us to specify an x-axis and a y-axis.
```{r}
homerange %>% 
  filter(family=="salmonidae") %>%
  select(common.name, log10.mass) %>% 
  ggplot(aes(y=common.name, x=log10.mass))+ #notice the switch in x and y
  geom_col()
```

`geom_bar()` with `stat="identity"`
`stat="identity"` allows us to map a variable to the y-axis so that we aren't restricted to counts.
```{r}
homerange %>% 
  filter(family=="salmonidae") %>% 
  ggplot(aes(x=common.name, y=log10.mass))+
  geom_bar(stat="identity")
```

## Practice
1. Filter the `homerange` data to include `mammals` only.
```{r}
homerange %>% 
  filter(taxon=="mammals")
```

2. Are there more herbivores or carnivores in mammals? Make a bar plot that shows their relative numbers.
```{r}
homerange %>% 
  filter(class=="mammalia") %>%
  count(trophic.guild) %>% 
  ggplot(aes(x=trophic.guild, y=n))+
    geom_col()
```

3. Make a bar plot that shows the masses of the top 10 smallest mammals.
```{r}
homerange %>% 
  filter(class=="mammalia") %>% 
  top_n(-10, log10.mass) %>% 
  ggplot(aes(x=common.name, y=log10.mass))+
  geom_col()+
  coord_flip()
```
## Review
Now that you have been introduced to `ggplot`, let's practice a few more plot types. Remember that plots are built in layers: **plot= data + geom_ + aesthetics**. We have to specify each of these in order for a plot to be produced. If you get stuck, it is often helpful to stop and make a quick sketch of what you want or expect to see on a piece of scratch paper.  

Let's review using the `penguins` data. First, get an idea of the structure: Are the data tidy? Are there NA's? Are the variables discrete, categorical, or continuous?
```{r}
#penguins
```

In a previous lab, we asked how many penguins were measured on each island.
```{r}
#penguins %>% 
  #count(island)
```

Make this output more visual by adding a plot...
```{r}
#penguins %>% 
  #count(island) %>% 
  #ggplot(aes(x=island, y=n))+
  geom_col()
```

## Practice
1. What if we wanted a plot that showed the number of measured penguins for each species?
```{r}
#penguins %>% 
  #count(species) %>% 
  #ggplot(aes(x=species, y=n))+
  geom_col()
```

2. How about average bill length by sex?
```{r}
#penguins %>% 
 # filter(sex!="NA") %>% 
  #group_by(sex) %>% 
  #summarize(ave_bill_length=mean(bill_length_mm)) %>% 
  #ggplot(aes(x=sex, y=ave_bill_length))+
  #geom_col()
```


## Box Plots
For the next series of examples, we will use the `homerange` data.  **Database of vertebrate home range sizes.**   

Reference: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
Data: http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1  

```{r message=FALSE, warning=FALSE}
homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")
```

Boxplots help us visualize a range of values. So, on the x-axis we typically have something categorical and the y-axis is the range. In the case below, we are plotting `log10.mass` by taxonomic class in the `homerange` data. `geom_boxplot()` is the geom type for a standard box plot. The center line in each box represents the median, not the mean.

Let's look at the variable `log10.mass` grouped by taxonomic class.
```{r}
homerange %>% 
  group_by(class) %>% 
  summarize(min_log10.mass=min(log10.mass),
            max_log10.mass=max(log10.mass),
            median_log10.mass=median(log10.mass))
```

```{r}
homerange %>% 
  ggplot(aes(x = class, y = log10.mass)) +
  geom_boxplot()
```

## Practice
1. There are more herbivores than carnivores in the homerange data, but how do their masses compare? Make a summary and boxplot that compares their masses. Use `log10.mass`.
```{r}
homerange %>%
  ggplot(aes(x = trophic.guild, y = log10.mass)) +
  geom_boxplot()
```

2. Have a closer look at carnivorous mammals. Summarize the range of log10.mass by family.
```{r}
homerange %>%
  filter(trophic.guild=="carnivore" & class=="mammalia") %>% 
  select(family, trophic.guild, log10.mass) %>%
  ggplot(aes(x=family, y=log10.mass))+
  geom_boxplot()+
  coord_flip()
```

## Load the data
Let's revisit the mammal life history data to practice our ggplot skills. The [data](http://esapubs.org/archive/ecol/E084/093/) are from: *S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.*
```{r message=FALSE, warning=FALSE}
#life_history <- read_csv("data/mammal_lifehistories_v2.csv", na="-999") %>% clean_names()
```

## Bar Plots
Bar plots count the number of observations in a categorical variable. What is the difference between `geom_bar` and `geom_col`? Make two bar plots showing the number of observations for each order using each geom type.

`geom_col`
```{r}
#life_history %>% 
  #count(order, sort=T) %>% #this is the same as arrange
  #ggplot(aes(x=order, y=n))+
  geom_col()+
  coord_flip()
```

`geom_bar`
```{r}
life_history %>%
  ggplot(aes(x=order))+
  geom_bar()+
  coord_flip()
```

Remember that ggplot build plots in layers. These layers can significantly improve the appearance of the plot. What if we wanted a bar plot of the mean mass for each order? Would we use `geom_bar` or `geom_col`?  
```{r}
life_history %>% 
  group_by(order) %>% 
  summarize(mean_mass=mean(mass, na.rm=T)) %>% 
  ggplot(aes(x=order, y=mean_mass))+
  geom_col()+
  coord_flip()
```

There are a few problems here. First, the y-axis is in scientific notation. We can fix this by adjusting the options for the session.
```{r}
options(scipen=999)#cancels scientific notation for the session
```

Next, the y-axis is not on a log scale. We can fix this by adding `scale_y_log10()`.
```{r}
life_history %>% 
  group_by(order) %>% 
  summarize(mean_mass=mean(mass, na.rm=T)) %>% 
  ggplot(aes(x=order, y=mean_mass))+
  geom_col()+
  coord_flip()+
  scale_y_log10()
```

Lastly, we can adjust the x-axis labels to make them more readable. We do this using `reorder`.
```{r}
life_history %>% 
  group_by(order) %>% 
  summarize(mean_mass=mean(mass, na.rm=T)) %>% 
  ggplot(aes(x=reorder(order, mean_mass), y=mean_mass))+
  geom_col()+
  coord_flip()+
  scale_y_log10()
```

## Scatterplots
Scatter plots allow for comparisons of two continuous variables. Make a scatterplot below that compares gestation time and weaning mass.
```{r}
life_history %>% 
  ggplot(aes(x=gestation, y=wean_mass))+
  geom_jitter(na.rm=T)+ #prevents overplotting
  scale_y_log10()
```

## Boxplots
Box plots help us visualize a range of values. So, on the x-axis we typically have something categorical and the y-axis is the range. Let's make a box plot that compares mass across taxonomic orders.
```{r}
life_history %>% 
  ggplot(aes(x=order, y=mass))+
  geom_boxplot(na.rm=T)+
  coord_flip()+
  scale_y_log10()
```

```{r}

```

## Aesthetics: Labels
Now that we have practiced scatter plots, bar plots, and box plots we need to learn how to adjust their appearance to suit our needs. Let's start with labeling x and y axes.  

For this exercise, let's use the `ElephantsMF` data. These data are from Phyllis Lee, Stirling University, and are related to Lee, P., et al. (2013), "Enduring consequences of early experiences: 40-year effects on survival and success among African elephants (Loxodonta africana)," Biology Letters, 9: 20130011. [kaggle](https://www.kaggle.com/mostafaelseidy/elephantsmf).
```{r}
elephants <- read_csv("data/elephantsMF.csv") %>% clean_names()
```

Make a plot that compares age and height of elephants.
```{r}
elephants %>% 
  ggplot(aes(x=age, y=height))+
  geom_point()+
  geom_smooth(method=lm, se=F)
```

The plot looks clean, but it is incomplete. A reader unfamiliar with the data might have a difficult time interpreting the labels. To add custom labels, we use the `labs` command.
```{r}
elephants %>% 
  ggplot(aes(x=age, y=height)) + 
  geom_point() + 
  geom_smooth(method=lm, se=F)+
  labs(title="Elephant Age vs. Height",#adds a title
       x="Age (years)", 
       y="Height(cm)")
```

We can improve the plot further by adjusting the size and face of the text. We do this using `theme()`. The `rel()` option changes the relative size of the title to keep things consistent. Adding `hjust` allows control of title position.
```{r}
elephants %>% 
  ggplot(aes(x=age, y=height)) + 
  geom_point() + 
  geom_smooth(method=lm, se=F)+
  labs(title="Elephant Age vs. Height",#adds a title
       x="Age (years)", 
       y="Height(cm)")+
  theme(plot.title=element_text(size=rel(1.5), hjust=0.5)) # hjust 0 left, 1 right
```

## Other Aesthetics
There are lots of options for aesthetics. An aesthetic can be assigned to either numeric or categorical data. `fill` is a common grouping option; notice that an appropriate key is displayed when you use one of these options.
```{r}
elephants %>% 
  ggplot(aes(x=sex, fill=sex))+ #fill is a grouping option
  geom_bar()
```

`size` adjusts the size of points relative to a continuous variable.
```{r}
life_history %>% 
  ggplot(aes(x=gestation, y=log10(mass), size=mass))+
  geom_point(na.rm=T)
```

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(palmerpenguins)
```

## Removes scientific notation
```{r}
options(scipen=999) #cancels the use of scientific notation for the session
```

## Data
**Database of vertebrate home range sizes.**  
Reference: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
Data: http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1  
```{r}
homerange <- 
  read_csv("data/Tamburelloetal_HomeRangeDatabase.csv", na = c("", "NA", "\\"))
```

## A few more useful aesthetics
There are many options to create nice plots in ggplot. One useful trick is to store the plot as a new object and then experiment with geom's and aesthetics. Let's setup a plot that compares `log10.mass` and `log10.hra.` Notice that we are not specifying a geom.
```{r}
p <- homerange %>% 
  ggplot(aes(x= log10.mass, y= log10.hra))
```

Play with point size by adjusting the `size` argument.
```{r}
p + geom_point(size=1.25)
```

We can color the points by a categorical variable.
```{r}
p + geom_point(aes(color=thermoregulation), size=1.5) #for scatterplots, color is the same as fill in barplots
```

We can also map shapes to another categorical variable.
```{r}
p + geom_point(aes(color=thermoregulation, shape=thermoregulation), size=1.5)
```

## Barplots and multiple variables
At this point you should be comfortable building bar plots that show counts of observations using `geom_bar()`. Last time we explored the `fill` option as a way to bring color to the plot; i.e. we filled by the same variable that we were plotting. What happens when we fill by a different categorical variable?  
Let's start by counting how many observations we have in each taxonomic group.
```{r}
homerange %>% count(taxon, sort=T)
```

Now let's make a bar plot of these data.
```{r}
homerange %>%
  ggplot(aes(taxon))+
  geom_bar()+
  coord_flip()+
  labs(title="Observations by Taxon", 
       x="Taxonomic Group")
```

By specifying `fill=trophic.guild` we build a stacked bar plot that shows the proportion of a given taxonomic group that is an herbivore or carnivore.
```{r}
homerange %>%
  ggplot(aes(taxon, fill=trophic.guild))+
  geom_bar()+
  coord_flip()+
  labs(title="Observations by Taxon", 
       x="Taxonomic Group")
```

We can also have counts of each trophic guild within taxonomic group shown side-by-side by specifying `position="dodge"`.
```{r}
homerange %>%
  ggplot(aes(taxon, fill=trophic.guild))+
  geom_bar(position="dodge")+
  coord_flip()+
  labs(title="Observations by Taxon", 
       x="Taxonomic Group")
```

Here is the same plot oriented vertically.
```{r}
homerange %>%
  ggplot(aes(taxon, fill=trophic.guild))+
  geom_bar(position="dodge")+
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  labs(title="Observations by Taxon", 
       x="Taxonomic Group")
```

We can also scale all bars to a percentage.
```{r}
homerange %>% 
  ggplot(aes(x = taxon, fill = trophic.guild))+
  geom_bar(position = position_fill())+ 
  scale_y_continuous(labels = scales::percent)+
  coord_flip()
```

## Practice
For this practice, let's use the [palmerpenguins](https://allisonhorst.github.io/palmerpenguins/articles/intro.html) data.  

1. Make a bar plot that shows counts of individuals by island. Fill by species, and try both a stacked bar plot and `position="dodge"`.
```{r}
penguins %>%
  ggplot(aes(island, fill=species))+
  geom_bar(position="dodge")+
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  labs(title="Individuals by Island", 
       x= "Island", 
       y= "Count")
```

```{r}

```

2. Make another bar plot that shows the number of individuals by sex on each island?
```{r}
penguins %>%
  ggplot(aes(island, fill=sex))+
  geom_bar(position="dodge")+
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  labs(title="Individuals by Sex", 
       x= "Island", 
       y= "Count", 
       fill="Sex")
```

## Using `group`
In addition to `fill`, `group` is an aesthetic that accomplishes the same function but does not add color.

Here is a box plot that shows `log10.mass` by taxonomic `class.`
```{r}
homerange %>% 
  ggplot(aes(x = class, y = log10.mass)) +
  geom_boxplot()
```

I use `group` to make individual box plots for each taxon within class.
```{r}
homerange %>% 
  ggplot(aes(x = class, y = log10.mass, group = taxon)) +
  geom_boxplot()
```

I can also use `fill` to associate the different taxa with a color coded key.
```{r}
homerange %>% 
  ggplot(aes(x = class, y = log10.mass, group = taxon, fill=taxon)) +
  geom_boxplot(alpha=.4)
```

## Install `RColorBrewer`, `paletteer`, and `ggthemes`
`RColorBrewer` and `paletteer` are helpful in selecting palettes of color that work with your data. Please install these now, we will work with them in part 2 of today's lab.
```{r, eval=FALSE, echo=TRUE}
#install.packages("RColorBrewer")
#install.packages("paletteer")
#install.packages("ggthemes")
```

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(RColorBrewer)
library(paletteer)
library(janitor)
```

```{r}
options(scipen=999) #cancels the use of scientific notation for the session
```

## Data
For this tutorial, we will use two data sets.  

[Desert ecology](http://esapubs.org/archive/ecol/E090/118/). The data are from: S. K. Morgan Ernest, Thomas J. Valone, and James H. Brown. 2009. Long-term monitoring and experimental manipulation of a Chihuahuan Desert ecosystem near Portal, Arizona, USA. Ecology 90:1708.  
```{r message=FALSE, warning=FALSE}
deserts <- read_csv("data/surveys_complete.csv")
```

[Homerange](http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1). The data are from: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
```{r message=FALSE, warning=FALSE}
homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv", na = c("", "NA", "\\"))
```

## Review
1. Use the `homerange` data to make a boxplot that shows the range of `log10.mass` by taxonomic class.
```{r}
homerange %>% 
  ggplot(aes(x=class, y=log10.mass))+
  geom_boxplot()
```

2. Now, add a layer to this plot that fills color for each box by taxonomic class
```{r}
homerange %>% 
  ggplot(aes(x=class, y=log10.mass, fill=class))+
  geom_boxplot()
```

3. Use `group` to show the range of `log10.mass` by taxonomic class.
```{r}
homerange %>% 
  ggplot(aes(x=class, y=log10.mass, group=taxon))+
  geom_boxplot()
```

4. Use `fill` to show the range of `log10.mass` by taxonomic class with color codes.
```{r}
homerange %>% 
  ggplot(aes(x=class, y=log10.mass, fill=taxon))+
  geom_boxplot()
```

## Line plots
Line plots are great when you need to show changes over time. Here we look at the number of samples for species DM and DS over the years represented in the `deserts` data. This takes some careful thought- we want to know how sampling has changed over time for these two species.    

Let's start by making a clear x and y so we know what we are going to plot.
```{r}
deserts %>% 
  filter(species_id=="DM"| species_id=="DS") %>% 
  mutate(year==as.factor(year)) %>% 
  group_by(year, species_id) %>% 
  summarize(n=n(), .groups= 'keep') %>% 
  pivot_wider(names_from= species_id, values_from = n)
```

```{r}
deserts %>% 
  filter(species_id=="DM"| species_id=="DS") %>% 
  mutate(year==as.factor(year)) %>% 
  group_by(year, species_id) %>% 
  summarize(n=n(), .groups= 'keep') %>% 
  ggplot(aes(x=year, y=n, group=species_id, color=species_id))+
  geom_line()+
  geom_point(shape=2)+ #you can experiment eith shapes
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  labs(title="Number of Samples for Species DM & DS", 
       x="Year", 
       y="n")
```

## Histograms
Histograms are frequently used by biologists; they show the distribution of continuous variables. As students, you have seen histograms of grade distributions. A histogram `bins` the data and you specify the number of bins that encompass a range of observations. For something like grades, this is easy because the number of bins corresponds to the grades A-F. By default, R uses a formula to calculate the number of bins but some adjustment may be required.  

What does the distribution of body mass look like in the `homerange` data?
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass)) +
  geom_histogram(bins=20)+ #we can adjust the number of bins with the bins argument
  labs(title = "Distribution of Body Mass")
```

Let's play with the colors. This shows all 657 of R's built-in colors. Notice that `color` and `fill` do different things!
```{r}
#grDevices::colors()
```

Let's rebuild the histogram, but this time we will specify the color and fill. Do a little experimentation on your own with the different colors.
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass)) +
  geom_histogram(color = "black", fill = "thistle3", bins=10)+
  labs(title = "Distribution of Body Mass")
```

## Density plots  
Density plots are similar to histograms but they use a smoothing function to make the distribution more even and clean looking. They do not use bins.
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass)) +
  geom_density(fill="deepskyblue4", alpha  =0.4, color = "black")+ #alpha is the transparency
  labs(title = "Distribution of Body Mass")
```

I like to see both the histogram and the density curve so I often plot them together. Note that I assign the density plot a different color.
```{r message=FALSE}
homerange %>% 
  ggplot(aes(x=log10.mass)) +
  geom_histogram(aes(y = after_stat(density)), fill = "deepskyblue4", alpha = 0.4, color = "black")+
  geom_density(color = "red")+
  labs(title = "Distribution of Body Mass")
```

## Practice
1. Make a histogram of `log10.hra`. Make sure to add a title.
```{r}
homerange %>% 
  ggplot(aes(x = log10.hra)) +
  geom_histogram(color = "black", fill = "springgreen4", bins=10)+
  labs(title = "Distribution of Homerange")
```

2. Now plot the same variable using `geom_density()`.
```{r}
homerange %>% 
  ggplot(aes(x = log10.hra)) +
  geom_density(fill="springgreen4", alpha  =0.4, color = "black")+
  labs(title = "Distribution of Homerange")
```

3. Combine them both!
```{r}
homerange %>% 
  ggplot(aes(x=log10.hra)) +
  geom_histogram(aes(y = after_stat(density)), fill ="springgreen4", alpha = 0.4, color = "black")+
  geom_density(color = "tan3")+
  labs(title = "Distribution of Homerange")
```

## Create Categories with mutate and case_when()
`case_when()` is a very handy function from `dplyr` which allows us to calculate a new variable from other variables. We use `case_when()` within `mutate()` to do this.`case_when()` allows us to specify multiple conditions. Let's reclassify the body mass variable into a new factor variable with small, medium, and large animals. In this case, we are making a continuous variable into a categorical variable.  
```{r}
homerange %>% 
  select(log10.mass) %>% 
  summarize(min=min(log10.mass), 
            max=max(log10.mass))
```
```{r}
summary(homerange$log10.mass)
```


Here we check how the newly created body mass categories compare across `trophic.guild`. 
```{r}
homerange %>% 
  mutate(mass_category=case_when(log10.mass<=1.75 ~ "small", 
                                 log10.mass>1.75 &log10.mass<=2.75 ~ "medium", log10.mass>2.75 ~ "large")) %>% 
  ggplot(aes(x=mass_category, fill=trophic.guild))+
  geom_bar(position="dodge")
```

## Practice
1. Use case_when() to make a new column `range_category` that breaks down `log10.hra` into very small, small, medium, and large classes based on quartile.

```{r}
#install.packages("gtools")
library(gtools)
```
```{r}
quartiles <- quantcut(homerange$log10.hra)
table(quartiles)
```

```{r}
homerange %>% 
  mutate(range_category=case_when(log10.hra<3.65 ~ "very small", log10.hra>=3.65 & log10.hra<=4.59 ~ "small", log10.hra>4.59 &log10.hra<=6.02~ "medium", log10.hra>6.02 ~ "large"))
```

## ggplot themes
There are many options to change the theme of your plots within ggplot. Have a look [here]https://ggplot2.tidyverse.org/reference/ggtheme.html) for a list of the themes.  

Let's start by building a simple barplot.  
```{r}
p <- homerange %>% 
  ggplot(aes(x=taxon, fill=trophic.guild))+
  geom_bar(na.rm=T, position="dodge")
```

Have a look at the `linedraw` theme; I am adding it as another layer.
```{r}
p + theme_linedraw()+
  theme(axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```

## Practice
1. Build a scatterplot that compares `log10.mass` and `log10.preymass`. Color the points by `taxon`. Store this plot as object `q`.
```{r}
q <- homerange %>% 
  ggplot(aes(x=log10.mass, y=log10.preymass, color=taxon))+
  geom_point(na.rm=T)
q
```

2. Use plot `q` but add the classic theme.
```{r}
q+ theme_classic()
```

## Legends
There are lots of options to manipulate legends. Have a look [here](http://www.sthda.com/english/wiki/ggplot2-legend-easy-steps-to-change-the-position-and-the-appearance-of-a-graph-legend-in-r-software).
```{r}
p+theme_linedraw()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```

## Not enough? Try using `ggthemes`
There are many packages that include additional themes, one of which is [ggthemes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/). Some of these are nice because they are designed to mimic the look of popular publications.
```{r}
#install.packages("ggthemes")
library(ggthemes)
```

Here is a list of the `ggthemes`
```{r}
#ls("package:ggthemes")[grepl("theme_", ls("package:ggthemes"))]
```

```{r}
p + 
  theme_fivethirtyeight()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL, # removes label from x axis
       y= "n",
       fill= "Trophic Guild")
```

## Practice
1. Use plot `q` and try the `theme_economist()` theme.
```{r}
p + 
  theme_economist()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL, # removes label from x axis
       y= "n",
       fill= "Trophic Guild")
```

## RColorBrewer
The default colors used by ggplot are often uninspiring. They don't make plots pop out in presentations or publications, and you may want to use a customized palette to make things visually consistent.  

Access the help for `RColorBrewer`.
```{r}
?RColorBrewer
```

The thing to notice is that there are three different color palettes: 1) sequential, 2) diverging, and 3) qualitative. Within each of these there are several selections. You can bring up the colors by using `display.brewer.pal()`. Specify the number of colors that you want and the palette name.
```{r}
display.brewer.pal(5,"Blues") #sequential palette
```

The [R Color Brewer](http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3) website is very helpful for getting an idea of the color palettes. To make things easy, use these two guidelines:

+`scale_colour_brewer()` is for points  
+`scale_fill_brewer()` is for fills  

Here I chose the `Paired` palette. Take a moment and experiment with other options.
```{r}
p+scale_fill_brewer(palette = "Paired")+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```

## Practice
1. Keep using the `same `q` plot, but test out one of R Color Brewer palettes.
```{r}
q+scale_colour_brewer(palette = "PuBuGn")+
  theme(legend.position = "bottom")
```

## Manually Setting Colors
You can also use `paleteer` to build a custom palette for consistency. To access the `paleteer` collection, I add it to a new object.
```{r}
colors <- paletteer::palettes_d_names
```

Now we can display the palettes. Assign the palette to `my_palette` and then build this base R bar plot. There are a lot of options; `paleteer` is a collection of popular palettes. I really like the [`ggsci` package] (https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html)
```{r}
my_palette <- paletteer_d("ggsci::nrc_npg") #store your palette
```

```{r}
barplot(rep(1,6), axes=FALSE, col=my_palette)
```

Now we just identify `my_palette` as part of `scale_fill_manual()`
```{r}
p+scale_fill_manual(values=my_palette)+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```

## Practice
1. Keep using the `same `q` plot, but test out one of `paleteer` themes.
```{r}

```

## Faceting
[Faceting](https://ggplot2-book.org/facet.html) is one of the amazing features of ggplot. It allows us to make multi-panel plots for easy comparison.  Here is a boxplot that shows the range of log10.mass by taxon.
```{r}
homerange %>% 
  ggplot(aes(x=taxon, y=log10.mass))+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 60, hjust=1))
```

There are other categorical variables that might be interesting to overlay. `facet_wrap()` makes a ribbon of panels by a specified categorical variable and allows you to control how you want them arranged.  
```{r}
homerange %>% 
  ggplot(aes(x=taxon, y=log10.mass))+
  geom_boxplot()+
  facet_wrap(~trophic.guild)+
  theme(axis.text.x = element_text(angle = 60, hjust=1))
```

`facet_grid()` allows control over the faceted variable; it can be arranged in rows or columns. rows~columns. 
```{r}
homerange %>% 
  ggplot(aes(x=taxon, y=log10.mass))+
  geom_boxplot()+
  facet_grid(trophic.guild~.)+
  theme(axis.text.x = element_text(angle = 60, hjust=1))
```

`facet_grid()` will also allow the comparison of two categorical variables, just remember a~b where a is rows and b is columns.  
```{r}
homerange %>% 
  ggplot(aes(x=taxon, y=log10.mass))+
  geom_boxplot()+
  facet_grid(trophic.guild~thermoregulation)+
  theme(axis.text.x = element_text(angle = 60, hjust=1))
```
